Teaching (Digital) Epigraphy

### 2016-02-05 source/48/EAGLE16-lame-et-al-teaching.tex
altri paper sistemati secondo istruzioni

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Teaching (Digital) Epigraphy}
\author[ilc]{Marion Lamé}
\author[zag]{Ivan Radman-Livaja}
\author[mau]{Bruce Robertson}
\author[isti]{Federico Ponchio\corref{first}}
\address[ilc]{Istituto di Linguistica Computazionale}
\address[isti]{Visual Computing Laboratory, ISTI-CNR}
\address[zag]{Archaeological Museum in Zagreb}
\address[mau]{Mount Allison University}
\cortext[first]{Corresponding author. Email: federico.ponchio@isti.cnr.it}

@@

\maketitle
\begin{abstract}
Although having the experience of directly manipulating an ancient text-bearing artifact is important for developing the
skills of an epigrapher – an experience that books and photo cannot replace – such access to primary sources is often
problematic. In this article we present our experience with teaching students to transcribe and interpret Roman
inscribed lead tags, using a Digital Autoptic Process (DAP) in a web environment, so to develop basic competences in
epigraphy and digital epigraphy.
\end{abstract}
\keywords{Educational project, digital epigraphy, epigraphy, open access, primary sources}

@@

\noindent Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the \emph{Tesserarum Sisciae Sylloge} (TSS) through
a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

@@

\begin{figure}[!h]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img001.jpg}
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img002.jpg}
\caption{No. 12582.}
\label{fig:12582}
\end{figure}

A traditional print photograph, such as this one (No. 12582, Fig.~\ref{fig:12582}) drawn from the archives of the AMZ, limits the student’s
opportunity to learn to decipher these tags. Even if a professional photographer took this picture with a good quality
camera in a fully equipped photographic studio, the reading of the inscription stays riddled with uncertainties and a
researcher would have to check the tag \emph{de visu} in order to offer a plausible transcription.

@@

\end{figure}
Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like \emph{LANA, PAN(N)UM, TVNICA, SAGVM, P(A)ENVLA, PAL(L)A, PALLIOLUM, LODIX, BANATA}, and \emph{ABOLLA}, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like \emph{l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab} is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\citet{radman_livaja_segestica_2008}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014},  for analogous tags with \citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_2011}; \citet[121-137/93, 215-222]{schwinden_romerzeitliche_1985}; \citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.

@@

\noindent In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g., No. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: \url{http://www.labcd.unipi.it/laboratorio}.}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\citet{gorman_opening_2015}
writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.''}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in \citet{Lame2015}, on a dispositive analysis
and the three fundamental systems of an inscription: writing system (wSystem), textual system (tSystem) and contextual
system (cSystem). Three tools compose the DAP: RTI web viewer, TSS viewer and MarkOut tool. The last two were
specifically designed for the digital edition of the corpus TSS. The following paragraphs present those three DAP
tools. Due to the material properties of the lead tags, it is impossible to capture the appearance of the letters with
a single image. When holding a tag it is necessary to change the light angle to enhance the different marks on the
surface. RTI is a computational photography technique that enables the interactive re-lighting of the observed object
from any light direction. This is accomplished combining into a single compact data structure a set of fixed camera
photographs taken under many different illumination conditions. The acquisition process is inexpensive and does not
require labor intensive steps, unlike 3d model acquisition. RTI visualization allows the user to virtually ‘tilt’ the
object and recreate different lighting conditions, mimicking part of the autoptic process. Another advantage is the
magnification provided by the high resolution camera employed. The widespread adoption in browsers of WebGL (Khronos
group 2009) recently enable RTI visualization on the web. We used the RTI web viewer developed by Visual Computing Lab,
ISTI-CNR. High resolution re-illuminable images allow the selection of the best view for each mark, thus allowing the
user to obtain an adequate readability of the lead tags through an intuitive interface. Another important advantage of
having the tags available online is that the students can easily compare a tag with any other tag already in the
database and compare occurrences of a letter in other already transcribed tags. Finally, the Web access facilitates the
discussion amongst students and experts worldwide.

@@

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.10]{EAGLE16lameetalteaching-img003.png}
\includegraphics[scale=0.10]{EAGLE16lameetalteaching-img004.png}
\caption{MarkOut screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
Latin characters with the position and the shape of the writing: U+0045 with II (identified as E2 by students) and
U+004C with L (L1 or L5).}
\label{fig:markout}
\end{figure}

@@

\begin{minipage}[t]{0.25\textwidth}
\textbf{Obverse}\\
At\d{e}\d{r}\\
ivs
\end{minipage}
\begin{minipage}[t]{0.25\textwidth}
(palimpsest)\\
\d{S}\d{i}\d{i}\d{x}\d{t}\d{i}\\
i onis n\\
i . n i 
\end{minipage}
\begin{minipage}[c]{0.30\textwidth}
%\begin{figure}
%\centering
\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img005a.png}
%\end{figure}
\end{minipage}

\begin{minipage}[t]{0.25\textwidth}
\textbf{Reverse}\\
la(na) p(ondo) i\\
cor(ticea)\\
\sout{X} =-\pounds\\
\end{minipage}
\begin{minipage}[t]{0.25\textwidth}
(palimpsest)\\
. . . s\\
. . v\\
  iir 
\end{minipage}
\begin{minipage}[c]{0.30\textwidth}
%\begin{figure}
%\centering
\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img005b.png}
%\end{figure}
\end{minipage}

@@

\section{Practicing Epigraphy}

\noindent Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
transcription and drawings) and to classify the written signs according to the following chart. Such preliminary
classification gives a first idea of the main general shapes one can find on a lead tag. It is not a paleographical
study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering.

\subsection{Drawing With Help (No. 12563)}

\begin{minipage}[t]{0.18\textwidth}
\textbf{Obverse}\\
Aeli(i)\\
Tasti
\end{minipage}
\begin{minipage}[c]{0.27\textwidth}
%\begin{figure}
%\centering
\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
%\end{figure}
\end{minipage}
\begin{minipage}[t]{0.18\textwidth}
\textbf{Reverse}\\
\d{l}odicem\\
murtiola\d{m}\\
pondo vi s\\
\sout{X} vi s\\
\end{minipage}
\begin{minipage}[c]{0.25\textwidth}
%\begin{figure}
%\centering
\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img007.png}
%\end{figure}
\end{minipage}

No. 12563 is a rectangular lead tag of irregular form with a badly damaged and scratched surface. Nevertheless, the most
recent inscription is clearly readable while one may still discern traces of older inscriptions. An older inscription
might be visible on the obverse, perhaps series of numbers but its meaning remains unclear. Students have
transcriptions and drawings.

\begin{table}
{\small
\addtolength{\tabcolsep}{-0.7mm}
\begin{tabular*}{\textwidth}{llllllllll}
\toprule
\textbf{12563 (obv)} & \textbf{A} & \textbf{E} & \textbf{L} & \textbf{I} & \textbf{T} & \textbf{A} & \textbf{S} & \textbf{T} & \textbf{I}\\
\midrule
\textbf{Stud. 1} & \textit{A10} & \textit{E2} & \textit{L11} & \textit{I1} & \textit{T1} & \textit{A10} & \textit{S3} & \textit{T2} & \textit{I1} \\
\textbf{Stud. 2} & \textit{A10} & \textit{E2} & \textit{L5} & \textit{I1} & \textit{T2} & \textit{A10} & \textit{S4?} & \textit{T2} & \textit{I1}  \\
\bottomrule
\end{tabular*}

@@

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img009.png}
\caption{12563 Obverse. The two students' reconstructions}
\label{fig:1}
\end{figure}

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img008.png}
\caption{12563 Reverse. The two students' reconstructions}
\label{fig:2}
\end{figure}

\subsection{Drawing \& Deciphering Without Help (obv. No. 13053)}

@@

\begin{minipage}[t]{0.3\textwidth}
\textbf{Vale}\\
rius lis\\
Martia\\
\end{minipage}
\begin{minipage}[t]{0.3\textwidth}
(palimpsest)\\
\sout{X} \d{v}i\\
tiir \d{r} i\\
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img011a.png}
\end{minipage}

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img010.png}
\caption{13053 Obverse. Students' reconstructions}
\label{fig:3}
\end{figure}

@@

The last line, MARTIA, gave the student a hard time. Particularly the tail of the damaged S (end of the second line)
generates noise on the last letters.

@@

\subsection{Palimpsest (obv. No. 13053)}

The older, erased layers, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
g/t relationship between wSystem and tSystem was asked to students.

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img012.png}
\caption{12563 Obverse (palimpsest). The two students' reconstructions}
\label{fig:4}
\end{figure}

@@

\noindent From the educational standpoint, this pilot project succeeded tremendously. The students who undertook this
transcription in place of the usual research essay were, on average, far more motivated than their peers, and took
great pleasure and pride in their work. One student, whose cumulative grade at university to this point was below
average, excelled in this project and said that this was what he imagined university studies were going to be like. Two
others prepared a presentation based on this effort, and it was accepted as a 15 minute lecture at the university’s
annual undergraduate research seminar.

@@

\section*{Aknowledgment}
\noindent We wish to acknowledge the research assistance of T. M. Crandall,
History student at Mount Allison University (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
``La lettre romaine, entre inscriptions monumentales et écritures communes''. The following
students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Teaching (Digital) Epigraphy}
\author[ilc]{Marion Lamé}\corref{first}
\author[zag]{Ivan Radman-Livaja}
\author[mau]{Bruce Robertson}
\author[isti]{Federico Ponchio}
\address[ilc]{Centre Camille Jullian, MMSH, CNRS and Laboratorio di Cultura Digitale, University of Pisa}
\address[isti]{Visual Computing Laboratory, ISTI-CNR}
\address[zag]{Archaeological Museum in Zagreb}
\address[mau]{Mount Allison University}
\cortext[first]{Corresponding author. Email: mlame@mmsh.univ-aix.fr}

@@

\maketitle
\begin{abstract}
Although having the experience of directly manipulating an ancient text-bearing artifact is important for developing the
skills of an epigrapher – an experience that books and photo cannot replace – such access to primary sources is often
problematic. In this article we present our experience with teaching students to transcribe and interpret Roman
inscribed lead tags, using a Digital Autoptic Process (DAP) in a Web environment, so to develop basic competences in
epigraphy and digital epigraphy.
\end{abstract}
\keywords{educational project, digital epigraphy, epigraphy, Open Access, primary sources}

@@

\noindent Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
Scholarly \& Technical Experiment in Digital Epigraphy: the Study Case of the \emph{Tesserarum Sisciae Sylloge} (TSS) through
a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

@@

\begin{figure}[!h]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img002.jpg}
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img001.jpg}
\caption{No. 12582, obverse and reverse.}
\label{fig:12582}
\end{figure}

 A traditional print photograph, such as Fig.~\ref{fig:12582} (No. 12582), drawn from the archives of the AMZ, limits the student’s
opportunity to learn to decipher these tags. Even if a professional photographer took this picture with a good quality
camera in a fully equipped photographic studio, the reading of the inscription stays riddled with uncertainties and a
researcher would have to check the tag \emph{de visu} in order to offer a plausible transcription.

@@

\end{figure}
Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like \emph{LANA, PAN(N)UM, TVNICA, SAGVM, P(A)ENVLA, PAL(L)A, PALLIOLUM, LODIX, BANATA}, and \emph{ABOLLA}, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like \emph{L, LA, LAN, PAN, T, SAG, PAENV, PAL, LO, LOD, LODI, BANA, AB} is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\citet{radman_livaja_segestica_2008}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014}, for analogous tags see:\citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_2011}; \citet[121-137]{schwinden_romerzeitliche_1985}; \citet[215-222]{schwinden_zwei_1993};\citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.

@@

\noindent In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g.: Fig.~\ref{fig:12582}, no. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa, applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: \url{http://www.labcd.unipi.it/laboratorio}.}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\citet{gorman_opening_2015}
writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.''}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in \citet{lame_primary_2014}, on a dispositive analysis
and the three fundamental systems of an inscription: writing system (wSystem), textual system (tSystem) and contextual
system (cSystem). Three tools compose the DAP: RTI Web viewer, TSS viewer and MarkOut tool. The last two were
specifically designed for the digital edition of the corpus TSS. The following paragraphs present those three DAP
tools. Due to the material properties of the lead tags, it is impossible to capture the appearance of the letters with
a single image. When holding a tag it is necessary to change the light angle to enhance the different marks on the
surface. RTI is a computational photography technique that enables the interactive re-lighting of the observed object
from any light direction. This is accomplished combining into a single compact data structure a set of fixed camera
photographs taken under many different illumination conditions. The acquisition process is inexpensive and does not
require labor intensive steps, unlike 3d model acquisition. RTI visualization allows the user to virtually ‘tilt’ the
object and recreate different lighting conditions, mimicking part of the autoptic process. Another advantage is the
magnification provided by the high resolution camera employed. The widespread adoption in browsers of WebGL (Khronos
group 2009) recently enable RTI visualization on the Web. We used the RTI Web viewer developed by Visual Computing Lab,
ISTI-CNR. High resolution re-illuminable images allow the selection of the best view for each mark, thus allowing the
user to obtain an adequate readability of the lead tags through an intuitive interface. Another important advantage of
having the tags available online is that the students can easily compare a tag with any other tag already in the
database and compare occurrences of a letter in other already transcribed tags. Finally, the Web access facilitates the
discussion amongst students and experts worldwide.

@@

\begin{figure}[!hbp]
\centering
\includegraphics[scale=0.10]{EAGLE16lameetalteaching-img003.png}
\includegraphics[scale=0.10]{EAGLE16lameetalteaching-img004.png}
\caption{MarkOut screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
Latin characters with the position and the shape of the writing: U+0045 with II (identified as E2 by students) and
U+004C with L (L1 or L5).}
\label{fig:markout}
\end{figure}

@@

\begin{minipage}[t]{0.25\textwidth}
\textbf{Obverse}\\
At\d{e}\d{r}\\
ivs
\end{minipage}
\begin{minipage}[t]{0.25\textwidth}
(palimpsest)\\
\d{S}\d{i}\d{i}\d{x}\d{t}\d{i}\\
i onis n\\
i . n i 
\end{minipage}
\begin{minipage}[c]{0.30\textwidth}
%\begin{figure}
%\centering
%2.965cm,height=1.79cm
\includegraphics[width=\columnwidth]{EAGLE16lameetalteaching-img005a.png}
%\end{figure}
\end{minipage}

\begin{minipage}[t]{0.25\textwidth}
\textbf{Reverse}\\
la(na) p(ondo) i\\
cor(ticea)\\
\sout{X} =-\pounds\\
\end{minipage}
\begin{minipage}[t]{0.25\textwidth}
(palimpsest)\\
. . . s\\
. . v\\
  iir 
\end{minipage}
\begin{minipage}[c]{0.30\textwidth}
%\begin{figure}
%\centering
%2.845cm,height=1.896cm
\includegraphics[width=\columnwidth]{EAGLE16lameetalteaching-img005b.png}
%\end{figure}
\end{minipage}

@@

\newpage
\section{Practicing Epigraphy}

\noindent Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
transcription and drawings) and to classify the written signs according to the following chart. Such preliminary
classification gives a first idea of the main general shapes one can find on a lead tag. It is not a paleographical
study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering (see: Fig.~\ref{fig:table1}).

\begin{figure}[!hbp]
\centering
\includegraphics[width=\columnwidth]{4_glyphs.png}
\caption{Glyph Browser: quick access to all the label associated with the glyph's form.}
\label{fig:table1}
\end{figure}

\subsection{Drawing with help (No. 12563)}

\vspace{0,5cm}
\begin{minipage}[t]{0.18\textwidth}
\textbf{Obverse}\\
Aeli(i)\\
Tasti
\end{minipage}
\begin{minipage}[c]{0.27\textwidth}
%\begin{figure}
%\centering
%2.965cm,height=1.79cm
\includegraphics[width=\columnwidth]{EAGLE16lameetalteaching-img006.png}
%\end{figure}
\end{minipage}
\begin{minipage}[t]{0.18\textwidth}
\textbf{Reverse}\\
\d{l}odicem\\
murtiola\d{m}\\
pondo vi s\\
\sout{X} vi s\\
\end{minipage}
\begin{minipage}[c]{0.25\textwidth}
%\begin{figure}
%\centering
%2.845cm,height=1.896cm
\includegraphics[width=\columnwidth]{EAGLE16lameetalteaching-img007.png}
%\end{figure}
\end{minipage}

\vspace{0,5cm}

The tag no. 12563 is a rectangular lead tag of irregular form with a badly damaged and scratched surface. Nevertheless, the most
recent inscription is clearly readable while one may still discern traces of older inscriptions. An older inscription
might be visible on the obverse, perhaps series of numbers but its meaning remains unclear. Students have
transcriptions and drawings (see Fig.~\ref{fig:1} and Fig.~\ref{fig:2}).

\begin{table}[!hbp]
{\small
\addtolength{\tabcolsep}{-0.7mm}
\begin{tabular*}{\textwidth}{llllllllll}
\toprule
\textbf{12563 (obv)} & \textbf{A} & \textbf{E} & \textbf{L} & \textbf{I} & \textbf{T} & \textbf{A} & \textbf{S} & \textbf{T} & \textbf{I}\\
\midrule
\textbf{Stud. 1} & \textit{A10} & \textit{E2} & \textit{L1} & \textit{I1} & \textit{T1} & \textit{A10} & \textit{S3} & \textit{T2} & \textit{I1} \\
\textbf{Stud. 2} & \textit{A10} & \textit{E2} & \textit{L5} & \textit{I1} & \textit{T2} & \textit{A10} & \textit{S4?} & \textit{T2} & \textit{I1}  \\
\bottomrule
\end{tabular*}

@@

\begin{figure}[!hbp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img009.png}
\caption{No. 12563 Obverse: students' drawings of the most recent inscription.}
\label{fig:1}
\end{figure}

\begin{figure}[!hbp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img008.png}
\caption{No. 12563 Reverse: students' drawings of the most recent inscription.}
\label{fig:2}
\end{figure}

\subsection{Drawing \& Deciphering without help (obv. No. 13053)}

@@

\begin{minipage}[t]{0.3\textwidth}
Vale\\
rius lis\\
Martia\\
\end{minipage}
\begin{minipage}[t]{0.3\textwidth}
(palimpsest)\\
\sout{X} \d{v}i\\
tiir \d{r} i\\
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\includegraphics[scale=0.20]{EAGLE16lameetalteaching-img011a.png}
\end{minipage}

\begin{figure}[!hbp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img010.png}
\caption{No. 13053 Obverse: students' drawings of the most recent inscription.}
\label{fig:3}
\end{figure}

@@

The last line, MARTIA, gave the student a hard time. Particularly the tail of the damaged S (end of the second line)
generates noise on the last letters (see Fig.~\ref{fig:3}).

@@

The older, erased layers, on the obverse of tag No. 13053, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
g/t relationship between wSystem and tSystem was asked to students (see Fig.~\ref{fig:4}).

\begin{figure}[!hbp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img012.png}
\caption{No. 13053 Obverse: students' drawing of some palimpsest inscriptions.}
\label{fig:4}
\end{figure}

@@

\noindent From the educational standpoint, this pilot project succeeded tremendously. The Canadian students who undertook this
transcription in place of the usual research essay were, on average, far more motivated than their peers, and took
great pleasure and pride in their work. One student, whose cumulative grade at university to this point was below
average, excelled in this project and said that this was what he imagined university studies were going to be like. Two
others prepared a presentation based on this effort, and it was accepted as a 15 minute lecture at the university’s
annual undergraduate research seminar. Three students are currently working on an exercise of online edition and publication (\emph{Épigraphie en Réseau} - ISSN: 2108-7156) of a lead tag using the Markout file created (SVG).

@@

\section*{Aknowledgment}
\noindent We wish to acknowledge the research assistance of T. M. Crandall,
History student at Mount Allison University (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
``La lettre romaine, entre inscriptions monumentales et écritures communes''. The following
students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.
\nocite{Lame2015b}
\nocite{cherry_storytelling_2014}
\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}
